---
title: "Introduction to geomtextpath"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to geomtextpath}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(ggplot2)
library(geomtextpath)
```

## Text on a path

The goal of the geomtextpath package is to draw text on a curved path as an extension of the ggplot2 plotting library. Those already familiar with ggplot2 and the `geom_path()` and `geom_text()` geometries, can easily display a path or some text.

```{r path_and_text_spiral, fig.show = 'hold'}
t <- seq(5, -1, length.out = 1000) * pi
spiral <- data.frame(
  x = sin(t) * 1000:1,
  y = cos(t) * 1000:1
)
rhyme <- paste(
  "Like a circle in a spiral, like a wheel within a wheel,",
  "never ending or beginning on an ever spinning reel"
)

p <- ggplot(spiral, aes(x, y)) +
  coord_equal(xlim = c(-1000, 1000), ylim = c(-1000, 1000))
p + geom_path() + labs(subtitle = "geom_path()")
p + geom_text(
  data  = data.frame(x = 0, y = 0),
  size  = 4, label = rhyme
) + labs(subtitle = "geom_text()")
```

The `geom_pathtext()` extension follows these ggplot2 conventions and displays the path and the text by projecting the text along the curve.

```{r textpath_spiral}
p + geom_textpath(
  size  = 4, label = rhyme
) + labs(subtitle = "geom_textpath()")
```

